---
title: "Exercise 1 --- Basic Maps and Graphs in R"
subtitle: "BSTA 374 · 12 points · Due Thursday, September 10, 2026, 11:59 pm"
author: Danny Lim
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  eval: true
  echo: true
---

## What this exercise is

Your first graded work in this course. It covers the first two weeks: reading
spatial data with `sf`, understanding what a coordinate reference system is and
moving between them, and producing honest choropleth and point maps with both
`ggplot2` and `tmap`.

**No data to download.** Everything runs on the North Carolina SIDS data that
ships inside the `sf` package, so your analysis will reproduce on my machine
exactly as it does on yours --- which is the point.

Each part ends with **starting code**. Copy it, fill in the blanks marked
`______`, and adapt it. It is scaffolding, not a solution: the mechanical lines
are given so that your time goes into the parts that require judgement.

::: {.callout-important}
## Submission
Submit **one file** to Course Site:

- `LASTNAME_FI_Exercise1.qmd` --- the source

Example: `DELMELLE_EM_Exercise1.qmd`. Put your name in the document header.

**Due Thursday, September 10, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Before you submit: restart R (`Ctrl/Cmd + Shift + F10`) and render the document
from a clean session. If it does not run start to finish, it is incomplete ---
regardless of how good the maps are. This is the habit that matters most, and
we start enforcing it now.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(tmap)
 library(patchwork)
# add anything else you use -- and load it here, not halfway down

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

---

## Part A --- Read and describe the data (2 points)

Answer each in **one or two sentences of prose**, outside the code chunks, and
show the code you used to find out.

1. How many features does `nc` contain, and what does one row represent?

nrow(nc)
[1] 100

The dataset "nc" contains 100 features and one row represents each observation in the data.

2. What geometry type is stored? Why is it that type rather than `POLYGON`?
   *(Look at the coast.)*
  
The geometry type that the data is stored in is multi polygon and it is called a multipolygon rather than the polygon because the coasts has separated counties from the land mass

names(nc)
 [1] "AREA"      "PERIMETER" "CNTY_"     "CNTY_ID"   "NAME"      "FIPS"     
 [7] "FIPSNO"    "CRESS_ID"  "BIR74"     "SID74"     "NWBIR74"   "BIR79"    
[13] "SID79"     "NWBIR79"   "geometry" 


 st_crs(nc)$input
[1] "NAD27"



3. What coordinate reference system is the data in? Give its name, say whether it
   is geographic or projected, and say how you found out.
The coordinate reference system is NAD27 and it is geographic and unprojected and I found out through the code st_crs(nc)$input.

4. In your own words, what do `BIR74` and `SID74` measure? Why does having both
   matter more than having either one alone?

SID74 measures the number of sudden infant deaths and BIR74 is the number of births and having both is important to figure out the rate of sudden infant deaths.


### Starting code

```{r}
#| label: a-starter

# A1 -- how big is it, and what are the columns?
nrow(nc)
names(nc)
nc                       # the header block above the table says a lot

# A2 -- one geometry type per feature; count them
table(st_geometry_type(nc))

# A3 -- three ways of asking about the CRS
st_crs(nc)               # the full definition
st_crs(nc)$input         # the short name
st_is_longlat(nc)        # TRUE = geographic (degrees), FALSE = projected (metres)

# A4 -- the help page documents every column
?nc
```

---

## Part B --- Coordinate reference systems (2 points)

1. Transform `nc` to **EPSG:32119** (NAD83 / North Carolina State Plane, in
   metres) and store it as a **new object**. Do not overwrite `nc` --- you need
   both.
2. Plot the counties **before and after**, near enough to each other that a reader
   can compare them. One or two sentences on what visibly changed.
   
  The map looked more bent and had curves rather than be straight shaped. 
  
  
3. Compute the area of every county in **square kilometres**, using the projected
   version. Report the five largest.

Columbus 2450.349			
Robeson	 2439.238			
Sampson	 2437.410			
Bladen	 2288.424			
Wake.    2194.261	
   
4. Now compute areas from the *unprojected* object as well. Compare the two sets
   of numbers and explain, in two or three sentences, what `sf` did and why the
   projected version is the one you would report in a paper.


Unprojected :


Columbus	2450.831			
Robeson	  2439.935			
Sampson	  2438.121			
Bladen	  2289.053			
Wake	    2194.374	

The two sets of number are very similar to each other with only tiny decimal differences. I would use the projected version because the projected version is well suited for distances, areas, and buffers. Additionally, it is valid over the region it was designated for which is north carolina, making the projected version a good choice.


::: {.callout-note}
Step 4 will not give you an error, and the two sets of numbers will be close but
not identical. That is the interesting part --- the question is which *surface*
each answer refers to.
:::

### Starting code

```{r}
#| label: b-starter

# B1
nc_proj <- st_transform(nc, 32119)

# B2 -- one plot each
p1 <- ggplot(nc)      + geom_sf(fill = "grey90", colour = "white") + theme_void()
p2 <- ggplot(nc_proj) + geom_sf(fill = "grey90", colour = "white") + theme_void()

p1
p2
# side by side, if you like:  library(patchwork);  p1 + p2

# B3 -- st_area() returns square METRES, as a units object
nc_proj <- nc_proj |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc_proj), km^2)))

nc_proj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)

nc <- nc |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc), km^2)))

nc |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)


# B4 -- now do the same starting from nc (degrees) and compare
p1/p2
```

---

## Part C --- Choropleth maps (5 points)

This is the largest part of the exercise, and it is as much about judgement as
about code.

1. Add a column giving the **SIDS rate per 1,000 live births** for 1974--78.
2. Make a choropleth of the **raw counts** (`SID74`) and one of the **rate**,
   using `ggplot2`. Then write three or four sentences explaining why the two maps
   disagree, and which one you would put in front of a health department, and why.

The two maps disagree because the sudden infant death map only shows the number of SID cases while the rate74 map shows the risk because it divides the SID cases by the births. I would put the rate of infant deaths in front of the health department because it demonstrates a better picture and explains more than just showing sudden infant death cases. For example, if the sudden infant death cases were 10 in county A and 5 in county B, a SID map would show that county A has a bigger problem of SID cases than county B. However, if the live births of County A is 1000 and county B is 20, it shows that SID cases affect more people in county B proportionally, which shows the full picture.
   
3. Rebuild the rate map using **`tmap`**. Same information, other package.
4. Vary the **class intervals** on the rate map: produce it at least three ways
   --- for instance equal intervals, quantiles, and Jenks natural breaks. Two or
   three sentences on how much the apparent geography of risk changes. Name the
   classification you would defend, and say why.
I would defend quantiles and Jenks natural breaks over equal intervals because equal interval makes skewed data collapse into 1 class which is shown in the map by having many counties as similar color. Jenks natural breaks and quantiles better suited to be used because show the data distribution rather than trying to distribute the data equally like the equal intervals. Quantile and Jenks show the high risks county on the map which stands out, however the equal intervals did not show the same apparent geographical risk changes.
   
   
5. Every map you submit must have a **legend with units**, a **title**, thin
   boundaries that do not overwhelm the fill, and a **perceptually uniform,
   colour-blind-safe palette**. State in one sentence which palette you chose and
   why, including whether dark means high or low.
I chose the palette magma because it clearly shows dark points clearly indicating which counties have high SID rate cases and lighter colors which shows low SID rate cases making the map easy to interpret.



::: {.callout-tip}
## What earns full marks here
Not the prettiest map --- the map whose choices you can defend. A student who
picks quantiles and explains the trade-off in two clear sentences scores higher
than one who picks Jenks because it looked nicer.
:::

### Starting code

```{r}
#| label: c-starter

# C1 -- deaths per 1,000 live births
nc <- nc |>
  mutate(rate74 = 1000*(SID74/BIR74))

# C2 -- counts, then rates. Same code, one variable changed.
ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000") +
  labs(title = "rate of infant deaths per 1000") +
  theme_void()

ggplot(nc) +
  geom_sf(aes(fill = SID74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "SID Cases") +
  labs(title = "SID cases") +
  theme_void()

# other palettes to try:
#   scale_fill_viridis_c(option = "viridis" / "plasma" / "cividis" / "mako")
#   scale_fill_distiller(palette = "YlOrRd", direction = 1)
#   RColorBrewer::display.brewer.all()   # shows every ColorBrewer option

# C3 -- the same map in tmap
tmap_mode("plot")                        # "view" makes it interactive

tm_shape(nc) +
  tm_polygons(fill = "rate74") +
  tm_title("SID rate per 1000 people")

# C4 -- change ONLY the classification, so the comparison is fair
tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_title("Quantiles")

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_title("Jenks natural breaks")

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_title("equal intervals")

# then repeat with style = "equal" and style = "jenks"
# other options: "pretty", "fisher"
```

::: {.callout-warning}
## A note on `tmap` versions
`tmap` changed its argument names in version 4. If an example you find online
uses `tm_polygons(col = ...)` or `style = ...` directly and yours errors, check
`packageVersion("tmap")` and read the help page rather than guessing. The code
above is written for version 4.
:::

---

## Part D --- A point layer (2 points)

1. Look up the approximate longitude and latitude of **four cities in North
   Carolina** and build a small table holding the name and the two coordinates.

Coordinates W: -80.84
            N:  35.22

Source: Wikipediea
        Geonames
   
   Cite where you got the coordinates.
2. Turn that table into an `sf` object with the correct CRS. State which EPSG
   code you used and why.
I used EPSG:4267 which is for NAD27 which matches the st_crs(nc)$input, so that the layers of the city points and the county lines and polygons can line up correctly.
   
   
3. Plot the four cities **on top of** your rate choropleth, labelled legibly.
4. One sentence: what would go wrong if you built the `sf` object with the
   coordinate columns in the other order?
If we had the coordinate columns in the other order the points would be in the wrong locations and maybe not be close to North Carolina at all. 
### Starting code

```{r}
#| label: d-starter

# D1 -- North Carolina longitudes are NEGATIVE (west of Greenwich)
cities <- tibble(
  name = c("Charlotte", "Raleigh", "Greensboro", "Durham"),
  lon  = c(-80.84, -78.639, -79.792, -78.8999),    
  lat  = c(35.22, 35.772, 36.073 , 35.994)
)

# D2 -- coords takes the column names in x, y order: longitude FIRST
cities_sf <- cities |>
  st_as_sf(coords = c("lon", "lat"), crs = 4267)

# D3 -- layers draw in order, so points come after polygons
ggplot() +
  geom_sf(data = nc, aes(fill = rate74), colour = "white", linewidth = 0.15) +
  geom_sf(data = cities_sf, size = 2.5, colour = "black") +
  geom_sf_text(data = cities_sf, aes(label = name), size = 3, nudge_y = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1, name = "per 1,000") +
  theme_void()

# check your cities landed in North Carolina and not in the ocean
```

---

## Part E --- Reproducibility and AI use (1 point)

1. Your document renders from a clean session, with every package loaded
   explicitly. *(Check this last, after you have stopped editing.)*
2. End the document with an **AI use note**: which tools you used, what for, and
   at which stage --- planning, coding, debugging, interpreting, or writing. Two
   or three sentences.
I used claude for debugging when I needed to download patchwork and also it helped me to interpret what crs = meant, but did not generate any code for me.


Remember the course policy: AI use is unrestricted and costs you nothing, but you
must be able to explain every line you hand in. Disclosure is the only
requirement.

### Starting code

```{r}
#| label: e-starter

sessionInfo()
```

---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct inspection of the object; clear, accurate answers | 2 |
| **B** | Correct transformation and areas; comparison explained | 2 |
| **C** | Both packages working; counts-vs-rates argument; class intervals compared and a defended choice; cartographic standards met | 5 |
| **D** | Point layer built with the right CRS and correctly overlaid | 2 |
| **E** | Renders from a clean session; AI use note present | 1 |
| | **Total** | **12** |

Writing counts for roughly 10% of the mark within each part: the prose should be
in complete sentences and free of obvious errors. Every part asks for
interpretation as well as code --- a submission that is all code and no sentences
cannot score above about 60%.

---

## Getting help

- Office hours, or email me --- with the code and the exact error message, not a
  screenshot of the whole console.
- The `sf` documentation is genuinely good: `vignette("sf1", package = "sf")`.
- Moraga Chapter 2 covers most of Parts A--C.
- If you are stuck for more than about thirty minutes on something that feels
  mechanical, that is a signal to ask rather than to persevere.